Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-22 paper.tex
Update spruce example to use local alleles

\begin{figure}
\begin{center}
\includegraphics[]{figures/data-scaling}
\end{center}
\caption{Compression performance on simulated genotypes.
Comparison of total stored bytes for VCF data produced 
by subsets of a large simulation of French-Canadians.
Sizes for $10^6$ samples are shown on the right.
Also shown for reference is the size of genotype matrix 
when encoded as two bits per diploid genotype (2bit), as used 
in the PLINK binary format. Default compression settings were 
used in all cases.
\label{fig-data-storage}}
\end{figure}

One of the key benefits of Zarr is its cloud-native design, 
but it also works well on standard file systems, where 
arrays and chunks are stored hierarchically in directories
and files (storage as a single Zip archive is also supported,
and has certain advantages as discussed later).
To enable comparison with the existing file-based ecosystem
of tools, we focus on Zarr's file system chunk storage in a series of illustrative 
benchmarks in the following sections before presenting cloud-based benchmarks. 
We compare primarily with
VCF/BCF based workflows using \texttt{bcftools} because this
is the standard practice, used in the vast majority of cases.
We also compare with two representative recent specialised utilities;
see~\cite{danek2018gtc,zhang2023gbc} for further benchmarks of 
these and other tools.
Genozip~\cite{lan2020genozip,lan2021genozip} is a tool focused 
on compression performance, which uses a custom file format 
and a CLI to extract VCF as text with various filtering options.
Savvy~\cite{lefaive2021sparse} is an extension of BCF which 
takes advantage of sparsity in the genotype matrix as well
as using PBWT-based approaches for improved compression.
Savvy provides a CLI as well as a C++ API.
Our benchmarks are based on genotype data 
from subsets of a large and highly realistic 
simulation of French-Canadians~\cite{anderson2023on}
(see Methods for details on the dataset and benchmarking methodology).
Note that while simulations cannot capture 
all the subtleties of real data, the allele frequency
and population structure patterns in this dataset 
have been shown to closely follow 
observations~\cite{anderson2023on} and so it provides 
a reasonable and easily reproducible data point 
when comparing such methods.
The simulations only contain genotypes without any additional
high-entropy QC fields, which is unrealistic 
(see the Genomics England case-study
for benchmarks on a large human dataset that includes 
many such fields). 
Note, however, that such minimal, genotype-only data 
is something of a best-case scenario for specialised genotype
compression methods using row-wise storage.

Fig~\ref{fig-data-storage} shows compression performance 
on up to a million samples for chromosome 21, with 
the size of the genotype-matrix encoded as 1-bit per haploid
call included for reference.
% 2bit  1689.153258 G
% vcf   81.375831 G
% = 20.75 X, 4.8%
Gzip compressed VCF performs remarkably well, compressing 
the data to around 5\% of the 
minimal binary encoding of a biallelic genotype matrix
for 1 million samples. 
BCF provides a significant improvement in compression
performance over VCF (note the log-log scale). Genozip has 
superb compression, having far smaller file sizes that the 
other methods (although somewhat losing its advantage at 
larger sample sizes). Zarr and Savvy have 
very similar compression performance in this example (23.75GiB
and 21.25GiB for $10^6$ samples, respectively). When using
a chunk size of $10^4$ variants $\times$ $10^3$ samples, Zarr
storage is reduced to 22.07GiB for $10^6$ samples (data not shown;
see Figs~\ref{fig-compression-chunksize} and \ref{fig-compression-chunksize-finegrained}
for more analysis on the effect of chunk size on compression).
It is remarkable that the simple approach of compressing
two dimensional chunks of the genotype matrix 
using the Zstandard compressor~\citep{collet2021rfc} and the 
bit-shuffle filter from Blosc~\cite{alted2010modern} 
(see Methods for details) produces 
compression levels competitive with the highly specialised methods
used by Savvy.

@@

The data shown in Fig~\ref{fig-whole-matrix-compute} for Zarr and Savvy
is based on custom programs written using their respective APIs
to implement the \texttt{af-dist} operation. The Zarr program uses
the Zarr-Python package to iterate over the decoded chunks of the 
genotype matrix and classifies genotypes within a chunk using a 14 line Python
function, accelerated using the Numba JIT compiler~\cite{lam2015numba}.
The allele frequencies and genotype counts are then analysed to produce 
the final counts within the allele frequency bins with 9 lines of 
Python using NumPy~\cite{harris2020array} functions. Remarkably, this 
short and simple Python program is substantially faster than the 
equivalent compiled C using \texttt{htslib} APIs on BCF (7.3 hours
vs 20.6 hours for 1 million samples). 
% num_samples  num_sites      tool  user_time  sys_time     wall_time  total_time
% 21      1000000    7254858     savvy    8451.25     16.59   8471.293500     hdd 2.352178
% 10      1000000    7254858  bcftools   73939.32     46.09  74023.083678     hdd 20.551503
% 27      1000000    7254858      zarr   26227.92    187.22  31476.507443     hdd 7.337539
The fastest method is the 
C++ program written using the Savvy API. This would largely seem
to be due to Savvy's excellent genotype decoding performance
% zarr 1.2 GiB
% savvy 6.6 GiB
% zarr_nshf 2.7 GiB
% zarr_lz4_nshf 2.9 GiB
(up to 6.6GiB/s vs 1.2GiB/s for Zarr on this dataset;
Fig~\ref{fig-whole-matrix-decode}).
% sav / zarr ratio: [0.68777206 0.75870945 0.69865406 0.64644888 0.68131752
% 0.89460906]
%     sequence_length  num_samples  num_sites                tool         size
% 58         48129895      1000000    7254858                zarr    23.752762
% 50         48129895      1000000    7254858         10kx1k.zarr    22.068925
% 51         48129895      1000000    7254858                 bcf    51.749294
% 52         48129895      1000000    7254858             genozip    10.691384
% 53         48129895      1000000    7254858  lz4.noshuffle.zarr   214.868005
% 54         48129895      1000000    7254858      noshuffle.zarr    31.208149
% 55         48129895      1000000    7254858                 sav    21.249436
% 56         48129895      1000000    7254858                 tsk     1.802636
% 57         48129895      1000000    7254858                 vcf    81.375831
% 59         48129895      1000000    7254858            zarr.zip    24.240338
% 5          48129895      1000000    7254858                2bit  1689.153258
Turning off the BitShuffle filter for the Zarr dataset,
however, leads to a substantial increase in decoding speed
(2.7GiB/s) at the cost of a roughly 25\% increase in storage
space (31.2GiB up from 23.75GiB for 1 million samples; data not
shown). The LZ4 codec (again without BitShuffle) leads to slightly higher
decoding speed (2.9GiB/s), but at the cost of much larger storage requirements 
(214GiB; data not shown).
Given the relatively small contribution of genotypes to the
overall storage of real datasets (see the Genomics England example)
and the frequency that they are likely to be accessed, using 
Zstandard without BitShuffle 
would seem like a good tradeoff in many cases.
This ability to easily tune compression performance
and decoding speed on a field-by-field basis is a major strong
point of Zarr. The \texttt{vcf2zarr} utility also provides
functionality to aid with such storage schema tuning.

The discrepancy between elapsed time and total CPU time for Zarr 
benchmarks in
Figs~\ref{fig-whole-matrix-compute}--~\ref{fig-column-extract}
and~\ref{fig-whole-matrix-decode} is mostly due to the overhead
of accessing a large number of small chunk files and 
consequent reduced effectiveness of file-system read-ahead caching.
While having many independent chunks maps 
naturally to high-performance parallel access patterns
in the cloud store setting~\citep{durner2023exploiting} 
(see also the section on accessing Genomics England data stored on S3), 
such file access patterns can be problematic in the 
classical file-system setting.
The Zarr Zip store (which is simply a Zip archive of the standard directory 
hierarchy) provides a useful alternative here and can 
significantly reduce overall processing time. To illustrate this
we repeated the benchmarks in Fig~\ref{fig-whole-matrix-decode} in
which we measure the time required to decode the entire genotype
matrix (thus serving as a lower-bound for any subsequent computations).
In the case of $10^6$ samples, the Zip file requires slightly more 
% Full data table above ^
% 58         48129895      1000000    7254858                zarr    23.752762
% 59         48129895      1000000    7254858            zarr.zip    24.240338
storage space than the directory hierarchy (23.75GiB and 24.24G
for file system and Zip file, respectively)
and while total 
CPU time is similar (193m vs 187m), the elapsed wall clock
time for the Zip file version is significantly less 
(258m vs 188m).
        % tool  user_time  sys_time    cpu_time   wall_time
% 11      zarr   11383.82    191.01  192.913833  258.598778
% 29  zarr.zip   11198.37     43.13  187.358333  188.659252

Another advantage of Zarr is the availabilty of several 
independent implementations across different programming languages
(see Methods). Fig~\ref{fig-whole-matrix-compute-zarr-versions}
compares the CPU time required to perform the computations
of Fig~\ref{fig-whole-matrix-compute} 
for af-dist implementations using the Zarr-Python~\cite{zarrpython},
JZarr~\cite{jzarr},
and TensorStore~\citep{tensorstore} (C++ and Python) libraries.
The program using the Tensorstore C++ API is fastest, requiring 
about 20\% less CPU time than the version using Zarr-Python and Numba
used in our other benchmarks.

% num_samples  num_sites       tool  user_time  sys_time     wall_time
% storage  total_time
% 22      1000000    7254858       zarr   16709.55     38.14  16910.057917 hdd    4.652136
% 20      1000000    7254858     ts_cpp   13233.50     92.81  13573.456810 hdd    3.701753
% 21      1000000    7254858  zarr_java   43780.50    245.28  44134.462377 hdd   12.229383
% 23      1000000    7254858      ts_py   18027.43    642.40  19163.174182 hdd    5.186064
% Cpp / Zarr python [0.79197226]

@@

We illustrate this cost of row-wise encoding in
Fig~\ref{fig-subset-matrix-compute}, where we run the af-dist calculation
on a small fixed-size subset of the genotype matrices of
Fig~\ref{fig-data-storage}. The two-dimensional chunking of Zarr
means that this sub-matrix can be efficiently
extracted, and therefore the execution time depends very weakly on 
the overall dataset size, with the computation requiring around
2 seconds for 1 million samples. Because of their 
row-wise encoding, CPU time scales with the number of samples
for all the other methods.
Fig~\ref{fig-subset-matrix-compute-supplemental} shows performance
for the same operation when selecting half of the samples in the 
dataset.

\subsection{Extracting, inserting and updating fields}
\begin{figure}
\includegraphics{figures/column-extract}
\caption{Time to extract the genome position and write to a text file.
Total CPU time required to extract the POS field for BCF,
sav and Zarr formats
for the data in Figure~\ref{fig-data-storage}.
For the BCF file we used \texttt{bcftools query -f"\%POS$\backslash$n"}.
For sav, we used the Savvy C++ API to extract position for each variant
and output text using the \texttt{std::cout} stream. For Zarr, we read 
the variant\_position array into a NumPy array, and then wrote to
a text file using the Pandas \texttt{write\_csv} method. 
Zarr CPU time is dominated by writing the text output; we also show
the time required to populate a NumPy array with the data in Zarr,
which is 2 seconds. Wall-clock time (dotted line) is dominated 
in this case by file I/O. Note that our benchmarking methodology of 
dropping the file system caches before each run (see Methods) is
likely not reflective of real world usage where a small and 
frequently used field such as POS would remain in cache.
Time to output text for Savvy is not significant
for $> 1000$ samples (not shown).
\label{fig-column-extract}}
\end{figure}
We have focused on the genotype matrix up to this point, contrasting
Zarr with existing row-wise methods.
Real-world VCFs encapsulate much more than just the genotype
matrix, and can contain large numbers of additional fields.
Fig~\ref{fig-column-extract} shows the time required to extract 
the genomic position of each variant in the simulated benchmark 
dataset, which we can use as an indicative example of a per-variant 
query. Although Savvy is many times faster than \texttt{bcftools query}
here, the row-wise storage strategy that they share means that 
the entire dataset must be read into memory and 
decompressed to extract just one field from each record. Zarr
excels at these tasks: we only read and decompress the information required.

@@

As described in the Methods, conversion to Zarr using 
\texttt{vcf2zarr} is a two-step process. We 
first converted the 106 VCF files (12.81 TiB) for chromosome 2
into the intermediate columnar format (ICF). This task was 
split into 14,605 partitions, and distributed using the Genomics England
HPC cluster. The average run-time per partition was 20.7 min.
The ICF representation used a total
of 9.94 TiB over 3,960,177 data storage files. 
We then converted the ICF to Zarr, partitioned into
5989 independent jobs, with an 18.6 min average run time.
This produced a dataset with 44 arrays, consuming a 
total of 2.54 TiB of storage over 6,312,488
chunk files (using a chunk size of $10^4$ variants $\times 10^3$ 
samples). This is a roughly 5X reduction in total storage 
space over the original VCF. 
The top fields in terms 
of storage are detailed in Table~\ref{tab-genomics-england-data}.
We do not compare with other tools such as Genozip and 
Savvy here because  they have fundamental limitations 
(as shown in earlier simulation-based benchmarks),
and conversion of these large VCFs is a major undertaking.

@@

Finally, to explore the use of cloud computing services with VCF Zarr,
we uploaded the Zarr directory hierarchy for chromosome 2 to an AWS S3 
standard bucket. We repeated the benchmarks above on the same subset
of 562,640 variants, using the same 
single-threaded sequential Zarr code.
Extracting the position array required 15s CPU, 23.8s elapsed.
The af-dist calculation required 2.70min CPU, 5.23min elapsed. 
Creating the filter mask
required 3.05min CPU, 11.15min elapsed.

While these benchmarks are comparable with results obtained on the 
shared file-system in the Genomics England research environment,
the sequential single-threaded access patterns used do not map
well to modern cloud computing environments.
As each request to a cloud object store has a high latency, it 
is important to have many requests running concurrently to achieve 
good performance~\citep{durner2023exploiting}.
To explore the limits of performance we developed a simple prototype
Zarr implementation, designed to maximise concurrency using asynchrous
IO and multiple processes (see Methods for details).
Fig~\ref{fig-s3-throughput} shows the rate at which genotypes 
can be decoded to RAM as well as rate at which we can compute
af-dist (using Numba to classify genotypes, as before)
as we increase the number of parallel processes on
an EC2 c5.24xlarge instance with 96 vCPUs (48 physical).
The maximum decompression rate of 47.4GiB/s 
occurs at 64 processes, with a corresponding 
network data rate of 127 MB/s [FIXME confirm and convert to MiB/s].
Fig~\ref{fig-s3-network-throughput} explores the limitations imposed by
network access by measuring the rate at which we can download the chunks
into RAM (without decompressing) using the same parallel and asynchronous
architecture. Here, the maximum rate of 1.2GiB/s for a c5.9xlarge instance
and 2.4GiB/s for a c5n.9xlarge (high performance network) instance
was achieved with 32 processes. Genotype data can therefore be fetched 
from S3 around an order of magnitude more quickly than it can be decompressed 
into RAM. This has some interesting consequences
when we consider the goal of seeking the highest possible compression levels 
for genotype data and the usual cloud storage cost structuring models (see
Discussion).

Returning to Fig~\ref{fig-s3-throughput}, we can see that the af-dist
calculation is performed at a maximum rate of 25.5GiB/s using 128 
processes. At this rate, the af-dist calculation for the target region 
was completed in 4.12s [CHECKME] (compare to 5.23min for the naive single-threaded
method above).
This level of throughput would complete af-dist for all chromosomes for all samples
% compute: $4.08 * 1.27
% What are the numbers here??
% io: (((722000000/10000)*(79000/1000))/1000)*$0.0004
in 1.27 hours for a compute cost of \$5.18 and an IO cost of \$2.28.
(See \citep{durbin2020task,moore2021ome,gowan2022using} for further Zarr
benchmarks in cloud settings.)

@@

The compression was performed on the All of Us Researcher Workbench platform 
(based on Google Cloud), using a single compute node with 64 CPUs and 416 GiB 
of RAM, which cost \$3.80/hour. We first converted the VCF to an ICF, which 
took 6 hours and 17 minutes of effective runtime and utilized 343 hours of CPU 
time, achieving a parallel efficiency of 86\%. This step incurred a cost of 
\$23.88. The resulting ICF representation required 7.5 GiB of storage distributed 
across 91,232 files. Next, the ICF was converted to Zarr, which took 1 hour and 8 minutes, with 
a CPU time of 68 hours and 8 minutes and a parallel efficiency of 94\%, at a 
cost of \$4.31. This second Zarr representation occupied 8.5 GiB of storage across 
232,809 chunk files.

\begin{table}
\caption{Summary for a selection of the largest VCF Zarr columns produced 
    from the dataset using \texttt{vcf2zarr} with chunk sizes of 1,000 variants and 
    10,000 samples. For column details, see Table~\ref{tab-genomics-england-data}.
    \label{tab-top-fields-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=4.1]S[table-format=3.2]}
\toprule
{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
\midrule
/call_LAD & int16 & 2.26 GiB & 290.0 & 27.85% \\
/call_GQ & int8 & 1.89 GiB & 87.0 & 23.29% \\
/call_LA & int8 & 1.24 GiB & 260.0 & 15.28% \\
/call_genotype & int8 & 914.22 MiB & 370.0 & 11.00% \\
/call_RGQ & int16 & 729.57 MiB & 460.0 & 8.78% \\
/call_genotype_mask & bool & 606.02 MiB & 550.0 & 7.29% \\
/call_FT & object & 498.61 MiB & 2700.0 & 6.00% \\
\bottomrule
\end{tabular}
\end{table}

@@

Following the conversion process detailed in the previous section, we
first converted the 165 VCF files (7.33TiB) to ICF, followed by
conversion to Zarr. The tasks were run as single jobs on a compute
node with 2,048 GB RAM and 128 cores. Conversion to ICF required
13h45min runtime. The ICF representation used a total of 6.77 TiB over
1,342,569 data storage files. Conversion to Zarr required 18h51min
runtime, generating a dataset with 32 arrays, consuming a total of 6.6
TiB storage over 11,984,581 chunk files. The reduction in total
storage (1.1X) is negligible, mainly due to the inherent difficulty in
compressing the PL field. The top fields in terms of storage are
detailed in Table~\ref{tab-spruce-data}.

>>>>

\begin{figure}
\begin{center}
\includegraphics[]{figures/data-scaling}
\end{center}
\caption{Compression performance on simulated genotypes.
Comparison of total stored bytes for VCF data produced 
by subsets of a large simulation of French-Canadians.
Sizes for $10^6$ samples are shown on the right. Sizes 
for Savvy (21.25GiB) and Zarr (22.06GiB) are very similar.
Also shown for reference is the size of genotype matrix 
when encoded as two bits per diploid genotype (2bit), as used 
in the PLINK binary format.
\label{fig-data-storage}}
\end{figure}

One of the key benefits of Zarr is its cloud-native design, 
but it also works well on standard file systems, where 
arrays and chunks are stored hierarchically in directories
and files (storage as a single Zip archive is also supported).
To enable comparison with the existing file-based ecosystem
of tools, we focus on Zarr's file system chunk storage in a series of illustrative 
benchmarks in the following sections before presenting cloud-based benchmarks. 
We compare primarily with
VCF/BCF based workflows using \texttt{bcftools} because this
is the standard practice, used in the vast majority of cases.
We also compare with two representative recent specialised utilities;
see~\cite{danek2018gtc,zhang2023gbc} for further benchmarks of 
these and other tools.
Genozip~\cite{lan2020genozip,lan2021genozip} is a tool focused 
on compression performance, which uses a custom file format 
and a CLI to extract VCF as text with various filtering options.
Savvy~\cite{lefaive2021sparse} is an extension of BCF which 
takes advantage of sparsity in the genotype matrix as well
as using PBWT-based approaches for improved compression.
Savvy provides a CLI as well as a C++ API.
Our benchmarks are based on genotype data 
from subsets of a large and highly realistic 
simulation of French-Canadians~\cite{anderson2023on}
(see Methods for details on the dataset and benchmarking methodology).
Note that while simulations cannot capture 
all the subtleties of real data, the allele frequency
and population structure patterns in this dataset 
have been shown to closely follow 
observations~\cite{anderson2023on} and so it provides 
a reasonable and easily reproducible data point 
when comparing such methods.
The simulations only contain genotypes without any additional
high-entropy QC fields, which is unrealistic 
(see the Genomics England case-study
for benchmarks on a large human dataset that includes 
many such fields). 
Note, however, that such minimal, genotype-only data 
is something of a best-case scenario for specialised genotype
compression methods using row-wise storage.

Fig~\ref{fig-data-storage} shows compression performance 
on up to a million samples for chromosome 21, with 
the size of the genotype-matrix encoded as 1-bit per haploid
call included for reference.
% 2bit  1689.153258 G
% vcf   81.375831 G
% = 20.75 X, 4.8%
Gzip compressed VCF performs remarkably well, compressing 
the data to around 5\% of the 
minimal binary encoding of a biallelic genotype matrix
for 1 million samples. 
BCF provides a significant improvement in compression
performance over VCF (note the log-log scale). Genozip has 
superb compression, having far smaller file sizes that the 
other methods (although somewhat losing its advantage at 
larger sample sizes). Zarr and Savvy have 
almost identical compression performance in this example.
It is remarkable that the simple approach of compressing
two dimensional chunks of the genotype matrix 
using the Zstandard compressor~\citep{collet2021rfc} and the 
bit-shuffle filter from Blosc~\cite{alted2010modern} 
(see Methods for details) produces 
compression levels competitive with the highly specialised methods
used by Savvy.

@@

The data shown in Fig~\ref{fig-whole-matrix-compute} for Zarr and Savvy
is based on custom programs written using their respective APIs
to implement the \texttt{af-dist} operation. The Zarr program uses
the Zarr-Python package to iterate over the decoded chunks of the 
genotype matrix and classifies genotypes within a chunk using a 14 line Python
function, accelerated using the Numba JIT compiler~\cite{lam2015numba}.
The allele frequencies and genotype counts are then analysed to produce 
the final counts within the allele frequency bins with 9 lines of 
Python using NumPy~\cite{harris2020array} functions. Remarkably, this 
short and simple Python program is substantially faster than the 
equivalent compiled C using \texttt{htslib} APIs on BCF (6.9 hours
vs 20.6 hours for 1 million samples). 
% num_samples  num_sites      tool  user_time  sys_time     wall_time  total_time
%     1000000    7254858     savvy    8371.07      4.16   8377.718136  2.326453
%     1000000    7254858  bcftools   73939.32     46.09  74023.083678  20.551503
%     1000000    7254858      zarr   24709.64     29.32  24750.704171  6.871933
The fastest method is the 
C++ program written using the Savvy API. This would largely seem
to be due to Savvy's excellent genotype decoding performance
% zarr 1.2 GiB
% savvy 6.6 GiB
% zarr_nshf 3.9 GiB
(up to 6.6GiB/s vs 1.2GiB/s for Zarr on this dataset;
Fig~\ref{fig-whole-matrix-decode}).
% sequence_length  num_samples  num_sites       tool       size
% 40         48129895      1000000    7254858       zarr  22.071687
% 35         48129895      1000000    7254858        bcf  51.749294
% 36         48129895      1000000    7254858    genozip  10.691384
% 37         48129895      1000000    7254858        sav  21.249436
% 38         48129895      1000000    7254858        tsk   1.802636
% 39         48129895      1000000    7254858        vcf  81.375831
% 41         48129895      1000000    7254858  zarr_nshf  29.897540
Turning off the BitShuffle filter for the Zarr dataset,
however, leads to a substantial increase in decoding speed
(3.9GiB/s) at the cost of a roughly 25\% increase in storage
space (29.9GiB up from 22.1GiB for 1 million samples; data not
shown). The LZ4 codec (again without BitShuffle) leads to slightly higher
decoding speed, but at the cost of much larger storage requirements 
(219GiB; data not shown).
Given the relatively small contribution of genotypes to the
overall storage of real datasets (see the Genomics England example)
and the frequency that they are likely to be accessed, using 
Zstandard without BitShuffle 
would seem like a good tradeoff in many cases.
This ability to easily tune compression performance
and decoding speed on a field-by-field basis is a major strong
point of Zarr. The \texttt{vcf2zarr} utility also provides
functionality to aid with such storage schema tuning.

@@

We illustrate this cost of row-wise encoding in
Fig~\ref{fig-subset-matrix-compute}, where we run the af-dist calculation
on a small fixed-size subset of the genotype matrices of
Fig~\ref{fig-data-storage}. The two-dimensional chunking of Zarr
means that this sub-matrix can be efficiently
extracted, and therefore the execution time depends very weakly on 
the overall dataset size, with the computation requiring around
1 second for 1 million samples. Because of their 
row-wise encoding, CPU time scales with the number of samples
for all the other methods.
Fig~\ref{fig-subset-matrix-compute-supplemental} shows performance
for the same operation when selecting half of the samples in the 
dataset.

\subsection{Extracting, inserting and updating fields}
\begin{figure}
\includegraphics{figures/column-extract}
\caption{Time to extract the genome position and write to a text file.
Total CPU time required to extract the POS field for BCF,
sav and Zarr formats
for the data in Figure~\ref{fig-data-storage}.
For the BCF file we used \texttt{bcftools query -f"\%POS$\backslash$n"}.
For sav, we used the Savvy C++ API to extract position for each variant
and output text using the \texttt{std::cout} stream. For Zarr, we read 
the variant\_position array into a NumPy array, and then wrote to
a text file using the Pandas \texttt{write\_csv} method. 
Zarr CPU time is dominated by writing the text output; we also show
the time required to populate a NumPy array with the data in Zarr,
which is less than a second. Wall-clock time (dotted line) is dominated 
in this case by file I/O. Time to output text for Savvy is not significant
for $> 1000$ samples (not shown).
\label{fig-column-extract}}
\end{figure}
We have focused on the genotype matrix up to this point, contrasting
Zarr with existing row-wise methods.
Real-world VCFs encapsulate much more than just the genotype
matrix, and can contain large numbers of additional fields.
Fig~\ref{fig-column-extract} shows the time required to extract 
the genomic position of each variant in the simulated benchmark 
dataset, which we can use as an indicative example of a per-variant 
query. Although Savvy is many times faster than \texttt{bcftools query}
here, the row-wise storage strategy that they share means that 
the entire dataset must be read into memory and 
decompressed to extract just one field from each record. Zarr
excels at these tasks: we only read and decompress the information required.

@@

As described in the Methods, conversion to Zarr using 
\texttt{vcf2zarr} is a two-step process. We 
first converted the 106 VCF files (12.81 TiB) for chromosome 2
into the intermediate columnar format (ICF). This task was 
split into 14,605 partitions, and distributed using the Genomics England
HPC cluster. The average run-time per partition was 20.7 min.
The ICF representation used a total
of 9.94 TiB over 3,960,177 data storage files. 
We then converted the ICF to Zarr, partitioned into
5989 independent jobs, with an 18.6 min average run time.
This produced a dataset with 44 arrays, consuming a 
total of 2.54 TiB of storage over 6,312,488
chunk files. This is a roughly 5X reduction in total storage 
space over the original VCF. 
The top fields in terms 
of storage are detailed in Table~\ref{tab-genomics-england-data}.
We do not compare with other tools such as Genozip and 
Savvy here because  they have fundamental limitations 
(as shown in earlier simulation-based benchmarks),
and conversion of these large VCFs is a major undertaking.

@@

Finally, to explore the use of cloud computing services with VCF Zarr,
the zarr files for chromosome 2 were uploaded to an AWS S3 standard bucket.
Using a single-threaded process on an EC2 instance the benchmarks above were
repeated. Extracting the position array required 15s CPU, 23.8s elapsed.
The af-dist calculation required 2.70min CPU, 5.23min elapsed. Creating the filter mask
required 3.05min CPU, 11.15min elapsed. These times are comparable to the on-premises
HPC system above.

This single-threaded approach is not optimal for cloud computing, as to optimise
both runtime and costs parallel processing is required. Using 64 parallel processes
on a single c5.24xlarge EC2 instance with 96 vCPUs (48 physical) the zarr chunks of
genotype array could be retrived from S3 and decompressed at a rate of 50.8 GB/s.
The network data rate was 127 MB/s, far below the capbility of the instance,
indicating that this process was compute-constrained. When calculating af-dist after
decompression the throughput was 24.6 GB/s. With this parallel approach the 
af-dist calculation above was then completed in 4.12s elapsed time.
This level of throughput would complete af-dist for all chromosomes for all samples
% compute: $4.08 * 1.27
% io: (((722000000/10000)*(79000/1000))/1000)*$0.0004
in 1.27 hours for a compute cost of \$5.18 and an io cost of \$2.28.
(See \citep{durbin2020task,moore2021ome,gowan2022using} for further Zarr
benchmarks in cloud settings.)

@@

The compression was performed on a single compute node with 64 CPUs and 416 GiB of RAM.  
We first converted the VCF to an ICF, which took 6 hours and 17 minutes of  
effective runtime and utilized 343 hours of CPU time, achieving a parallel efficiency of 86\%.  
This step incurred a cost of \$23.88. The resulting ICF representation required 7.5 GiB of storage  
distributed across 91,232 files. Next, the ICF was converted to Zarr, which took 50 minutes and 32 
seconds of effective runtime and 49 hours and 49 minutes of CPU time, with a parallel efficiency of 92\%, 
at a cost of \$3.19. The final Zarr representation occupied 13 GiB of storage across 369,656 chunk files.

\begin{table}
\caption{Summary for a selection of the largest VCF Zarr columns produced 
the from the 245,388 exome-like samples from chromosome 20
using \texttt{vcf2zarr} and chunk sizes of 1,000 variants and 
10,000 samples.
For column details see the caption of Table~\ref{tab-genomics-england-data}.
\label{tab-all-of-us-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {type} & {storage} & {compress} & {\%total} \\
\midrule
/call\_GQ & int8 & 1.89G & 87.0 & 45.94\% \\
/call\_genotype & int8 & 914.22M & 370.0 & 21.70\% \\
/call\_RGQ & int16 & 729.57M & 460.0 & 17.32\% \\
/call\_genotype\_mask & bool & 606.02M & 550.0 & 14.39\% \\
/call\_genotype\_phased & bool & 17.17M & 9700.0 & 0.41\% \\
/variant\_allele & str & 4.74M & 110.0 & 0.11\% \\
/variant\_filter & bool & 2.87M & 0.9 & 0.07\% \\
% /variant\_AN & int32 & 908.75 K & 3.1 & 0.02\% \\
% /variant\_position & int32 & 810.77 K & 3.4 & 0.02\% \\
% /sample\_id & object & 357.15 K & 5.4 & 0.01\% \\
% /variant\_length & int16 & 152.4 K & 9.2 & 0.00\% \\
% /variant\_id & object & 56.18 K & 99.0 & 0.00\% \\
% /variant\_quality & float32 & 51.92 K & 54.0 & 0.00\% \\
% /variant\_id\_mask & bool & 51.89 K & 13.0 & 0.00\% \\
% /variant\_contig & int16 & 50.6 K & 28.0 & 0.00\% \\
% /region\_index & int32 & 13.93 K & 1.2 & 0.00\% \\
% /contig\_length & int64 & 9.24 K & 2.8 & 0.00\% \\
% /contig\_id & object & 8.33 K & 3.2 & 0.00\% \\
% /filter\_id & object & 4.48 K & 0.0 & 0.00\% \\
\bottomrule
\end{tabular}
\end{table}

@@

Following the conversion process detailed in the previous section, we
first converted the 165 VCF files (7.33TiB) to ICF, followed by
conversion to Zarr. The tasks were run as single jobs on a compute
node with 512 GB RAM and 128 cores using 120 cores. Conversion to ICF
required 13h45min runtime. The ICF representation used a total of 6.77
TiB over 1,340,804 data storage files. Conversion to Zarr required
18h36min runtime, generating a dataset with 32 arrays, consuming a
total of 6.6 TiB storage over 10,861,038 chunk files. The reduction in
total storage (1.1X) is negligible, mainly due to the inherent
difficulty in compressing the PL field. The output dataset storage
size is dominated by a major component, call\_PL, which accounts for
nearly 93\% of the total. call\_PL is yet another field with low
compression potential due to inherent noisiness.

Therefore, to reduce storage
requirements we generated a Zarr schema for encoding that uses local
allele fields with \texttt{vcf2zarr mkschema --local-alleles}, and
used the generated schema as input to \texttt{vcf2zarr encode}. Zarr conversion required 14h09min runtime, generating a dataset with 35 arrays, consuming a total of 2.3TiB storage over 11,235,556 chunk files, with a reduction in storage of 3.2X. Therefore, by using local allele fields we achieve approximately 3X improvement of compression. The top fields in terms of storage are detailed in Table~\ref{tab-spruce-data}.

\begin{table}
\caption{Summary for a selection of the largest VCF Zarr columns produced for 
1,063 VCFs on 12 spruce chromosomes using \texttt{vcf2zarr} with samples chunk size set to 1,063 , variants chunk size to 10,000, and using local allele fields to reduce output size. For column details see the caption of Table~\ref{tab-genomics-england-data}.
\label{tab-spruce-data}}

\begin{tabular}{llS[table-format=3.1]S[table-format=3.2]S[table-format=3.2]}
\toprule
Field & type & storage & compress & percentage \\
\midrule
/call_LPL & int16 & 1.26 TiB & 17.0 & 54.76% \\
/call_LA & int8 & 415.39 GiB & 18.0 & 17.63% \\
/call_genotype & int8 & 381.28 GiB & 19.0 & 16.18% \\
/call_genotype_mask & bool & 103.84 GiB & 71.0 & 4.41% \\
/variant_DP4 & int32 & 26.24 GiB & 2.1 & 1.11% \\
/variant_MQSBZ & float32 & 13.08 GiB & 1.1 & 0.56% \\
/variant_RPBZ & float32 & 13.04 GiB & 1.1 & 0.55% \\
/variant_MQBZ & float32 & 12.97 GiB & 1.1 & 0.55% \\
/variant_SCBZ & float32 & 12.94 GiB & 1.1 & 0.55% \\
/variant_VDB & float32 & 12.84 GiB & 1.1 & 0.55% \\
/variant_quality & float32 & 12.84 GiB & 1.1 & 0.55% \\
/variant_BQBZ & float32 & 12.64 GiB & 1.1 & 0.54% \\
/variant_SGB & float32 & 12.12 GiB & 1.2 & 0.51% \\
/variant_position & int32 & 9.65 GiB & 1.4 & 0.41% \\
/variant_AC & int16 & 7.42 GiB & 2.8 & 0.31% \\
/variant_DP & int32 & 6.71 GiB & 2.1 & 0.28% \\
/variant_allele & object & 5.37 GiB & 21.0 & 0.23% \\
/variant_AN & int16 & 3.05 GiB & 2.3 & 0.13% \\
/call_genotype_phased & bool & 1.75 GiB & 2100.0 & 0.07% \\
/variant_filter & bool & 1.46 GiB & 2.4 & 0.06% \\
/variant_MQ & int8 & 489.41 MiB & 7.3 & 0.02% \\
/variant_IMF & float32 & 160.95 MiB & 89.0 & 0.01% \\
/variant_IDV & int16 & 120.72 MiB & 59.0 & 0.01% \\
\bottomrule
\end{tabular}
\end{table}
